Towards a Pandemic-Scale Ancestral Recombination Graph of SARS-CoV-2

### 2023-04-14 paper.tex
Update analysis on exclusion criteria

\subsection{Recombinant Pango lineages}

In this section we examine the Long ARG to determine how well it reflects
the structure of previously identified recombinant lineages.
There are [high level summary of the pango recombinant lineages and samples
in the Long ARG].
long ARG (subsampled to mid-2022) contains 749 samples
from 33 Pango-designated recombinant lineages (XA to XAK; Supplementary Table)
[Probably could have a table here to summarise?]

% \item Total of 2078 recombinants.
% \item 749 samples from 33 Pango X lineage (recombinant) designations
% \item \alert{All but} 2 XP samples inferred to have recombinant ancestry
% (Almost certainly missing XP because of our alignment masking approach)

All but 2 XP samples inferred to have recombinant ancestry.

The sequences of 28 Pango recombinants trace back
to some recombinant node(s); however, the sequences of the five Pango
% jk: really? I thought it was just XP that wasn't inferred to be a
% recombinant.
recombinants XP, XU, XAA, XAG, and XAK do not.
A likely explanation for the XP
recombinant sequences not having a recombinant origin in the long ARG is that
its characteristic multibase deletion
(\url{https://github.com/cov-lineages/pango-designation/issues/481}) was masked
during preprocessing and therefore was not incorporated into the HMM.
% Eh?
Upon tracing the origin of XAG through the ARG, XAG might have been caused by
incorrect imputation of its parent nodes(?). XAK likely does not appear to have
a recombinant origin because of random subsampling, but it may with a more
complete tree.

[FIXME this isn't quite right - not all the sequences for a given pango
lineage trace back to the same recombination event, so won't have the same
intervals.]
The mosaic genome structure of all the 32 Pango recombinant lineages is concordant with
the structure determined by the Pango Lineage Designation Committee
(Supplementary Table 3). For example, the genome of XD (also known as
``Deltacron'') is inferred to consist of B.1.617.2 (Delta) on the 5’ end, BA.1.17
(Omicron) in the middle, and B.1.617.2 (Delta) on the 3’ end.

The wide ARG contains samples from three Pango-designated recombinant
lineages (XA, XB, and XC) and N samples, since it only includes data
up to mid-2021.
In the wide ARG, every sequence of the three Pango recombinants trace back to
some recombination node. XA, XB, and XC.

For the long ARG, we examine the lineages in increasing complexity of the
inferred evolutionary relationships between samples assigned to Pango
X lineages.

>>>>

\subsection{Recombinant Pango lineages}
[Intro bit saying what we're doing]

[Quick overview of the wide ARG.]
The wide ARG contains samples from three Pango-designated recombinant
lineages (XA, XB, and XC) and N samples, since it only includes data
up to mid-2021.
In the wide ARG, every sequence of the three Pango recombinants trace back to
some recombination node

[Details of the long ARG]
In this section we examine the Long ARG to determine how well it reflects
the structure of previously identified recombinant lineages.
There are [high level summary of the pango recombinant lineages and samples
in the Long ARG].
long ARG (subsampled to mid-2022) contains 749 samples
from 33 Pango-designated recombinant lineages (XA to XAK; Supplementary Table)

Of these 749 samples, all but 2 are inferred to descend from a recombination
node. These two samples are designated XP, and a likely explanation
for their not have an inferred recombinant origin in the long ARG is that
the characteristic multibase deletion for XP
(\url{https://github.com/cov-lineages/pango-designation/issues/481}) is masked
during preprocessing. An improved masking pipeline is an important avenue
for future development (see the Discussion).

38 of the remaining X designated samples descend from recombination
nodes that are ancestral only to that sample
(10 are XZ; 6 are XE;  3 each from
XN and XK; 2 each from XC, XS, XV, XQ and XAB; and 1 sample from each
of XB, XM, XJ, XAF, XAH and XAJ).
Such samples are likely to be enriched for sequencing errors and
lineage designation artefacts, and so we exclude them from further
analysis in this section.

The most recent recombination node for 79 samples
(XN: 53, XZ: 16, XAJ: 6, XE: 1, XAD: 1, XAH: 1, XAK: 1)
is the same. This node is ancestral to more than 127K samples,
and is likely to be a spuriously inferred recombination event.
Understanding the origin of this and other similar nodes, is
an important direction for future work (see Discussion).
For simplicity, we exclude these samples from further
analyses in this section.

We analyse the inferred evolutionary patterns of the
remaining 630 Pango X designated sequences in the following
subsections.

% The sequences of 28 Pango recombinants trace back
% to some recombinant node(s); however, the sequences of the five Pango
% % jk: really? I thought it was just XP that wasn't inferred to be a
% % recombinant.
% recombinants XP, XU, XAA, XAG, and XAK do not.
% A likely explanation for the XP
% recombinant sequences not having a recombinant origin in the long ARG is that
% its characteristic multibase deletion
% (\url{https://github.com/cov-lineages/pango-designation/issues/481}) was masked
% during preprocessing and therefore was not incorporated into the HMM.
% % Eh?
% Upon tracing the origin of XAG through the ARG, XAG might have been caused by
% incorrect imputation of its parent nodes(?). XAK likely does not appear to have
% a recombinant origin because of random subsampling, but it may with a more
% complete tree.

% [FIXME this isn't quite right - not all the sequences for a given pango
% lineage trace back to the same recombination event, so won't have the same
% intervals.]
% The mosaic genome structure of all the 32 Pango recombinant lineages is concordant with
% the structure determined by the Pango Lineage Designation Committee
% (Supplementary Table 3). For example, the genome of XD (also known as
% ``Deltacron'') is inferred to consist of B.1.617.2 (Delta) on the 5’ end, BA.1.17
% (Omicron) in the middle, and B.1.617.2 (Delta) on the 3’ end.

% For the long ARG, we examine the lineages in increasing complexity of the
% inferred evolutionary relationships between samples assigned to Pango
% X lineages.
